---
title: "Restricted Boltzmann Machine for Single-Cell Analysis"
author: "PreGraphModeling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Restricted Boltzmann Machine for Single-Cell Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6
)
```

## Introduction

This vignette demonstrates how to use the Restricted Boltzmann Machine (RBM) functionality in PreGraphModeling to model relationships between gene expression features and cell metadata. The RBM uses partial correlations estimated via pseudolikelihood to capture direct relationships between features while accounting for confounding effects.

## Key Features

- **Partial Correlation Estimation**: Compute partial correlations using pseudolikelihood for various families (ZINB, NB, Poisson, Gaussian)
- **RBM Fitting**: Connect gene expression features to metadata factors through learned weights
- **Parallel Processing**: Leverage `progressr` for efficient computation at scale
- **Visualization**: Create heatmaps of partial correlations and weights using ComplexHeatmap

## Installation

```{r eval=FALSE}
# Install from GitHub
devtools::install_github("GWMcElfresh/PreGraphModeling")

# Install BiocManager dependencies
BiocManager::install(c("ComplexHeatmap", "circlize"))

# Install suggested packages for full functionality
install.packages(c("progressr", "viridisLite"))
```

## Load Required Packages

```{r eval=FALSE}
library(PreGraphModeling)
library(Seurat)
library(ComplexHeatmap)
```

## Basic Workflow

The recommended workflow for RBM analysis has two steps:

1. **Pre-compute partial correlations** using `EstimatePartialCorrelationsFromSeurat()` - this allows you to inspect the correlations and reuse them across multiple RBM fits
2. **Fit the RBM** using `FitRBM()` with the pre-computed correlations

### 1. Estimate Partial Correlations

The first step is to compute partial correlations between features using pseudolikelihood estimation. This accounts for the zero-inflated nature of single-cell data.

```{r eval=FALSE}
# Load example Seurat object (e.g., PBMC dataset)
# For this example, we'll assume you have a Seurat object called 'pbmc'

# Recommended: Use the Seurat wrapper for convenience
pcor <- EstimatePartialCorrelationsFromSeurat(
  seuratObject = pbmc,
  family = "zinb",
  minNonZero = 10,
  parallel = TRUE,
  verbose = TRUE
)

PreGraphModeling:::.validate_partial_correlations(pcor_result)

# View the partial correlations object
print(pcor)

# Access the partial correlation matrix directly
pcor_matrix <- pcor$partial_cor
print(dim(pcor_matrix))

# The object contains:
# - partial_cor: the correlation matrix
# - features: valid features that passed filtering
# - family: the distribution family used
# - excluded_features: features excluded due to low counts
```

**Alternative:** You can also use the lower-level `EstimatePartialCorrelations()` function with a raw expression matrix if you need more control.

The function supports multiple distribution families:

- **"zinb"** (default): Zero-inflated negative binomial - best for single-cell count data
- **"nb"**: Negative binomial
- **"poisson"**: Poisson distribution
- **"gaussian"**: Gaussian (for normalized/continuous data)
- **"quasipoisson"**: Quasi-Poisson for overdispersed count data

### 2. Fit the RBM

Next, fit an RBM using the pre-computed partial correlations:

```{r eval=FALSE}
# Fit RBM with cell type as the hidden layer
# Pass the pre-computed partial correlations
rbm <- FitRBM(
  seuratObject = pbmc,
  hiddenFactors = "CellType",
  partialCorrelations = pcor, # Use pre-computed correlations
  cd_k = 5, # Contrastive Divergence steps
  n_epochs = 20, # Training epochs
  learning_rate = 0.01, # Step size
  parallel = TRUE,
  verbose = TRUE
)

# Print RBM summary
print(rbm)
summary(rbm)
```

**Why pre-compute?**

1. **Inspect before fitting**: You can examine the partial correlations before committing to RBM training
2. **Reuse across fits**: Fit multiple RBMs with different hidden factors without recomputing correlations
3. **Save computation time**: Partial correlation estimation is the most expensive step
4. **Save/load**: You can save partial correlations to disk and reload them later

```{r eval=FALSE}
# Reuse partial correlations for another RBM
rbm_multi <- FitRBM(
  seuratObject = pbmc,
  hiddenFactors = c("CellType", "Treatment"),
  partialCorrelations = pcor, # Same pre-computed correlations
  cd_k = 5,
  n_epochs = 20
)

# Save partial correlations for later use
saveRDS(pcor, "partial_correlations.rds")

# Load and reuse
pcor_loaded <- readRDS("partial_correlations.rds")
rbm_new <- FitRBM(pbmc, "CellType", partialCorrelations = pcor_loaded)
```

**Legacy mode:** If you don't provide `partialCorrelations`, FitRBM will compute them on-the-fly (backward compatible):

```{r eval=FALSE}
# This still works, but is not recommended for repeated fits
rbm_legacy <- FitRBM(
  seuratObject = pbmc,
  hiddenFactors = "CellType",
  family = "zinb", # Only used when computing on-the-fly
  parallel = TRUE
)
```

#### Contrastive Divergence Training

The RBM is trained using Contrastive Divergence (CD-k). Key parameters:

| Parameter | Default | Description |
|-----------|---------|-------------|
| `cd_k` | 1 | Number of Gibbs sampling steps (CD-1, CD-5, etc.) |
| `n_epochs` | 10 | Training iterations over the data |
| `learning_rate` | 0.01 | Step size for weight updates |
| `batch_size` | 100 | Mini-batch size |
| `momentum` | 0.5 | Momentum coefficient for updates |
| `weight_decay` | 0.0001 | L2 regularization |
| `persistent` | FALSE | Use Persistent CD (maintains chains) |

```{r eval=FALSE}
# More aggressive training with Persistent CD
rbm_trained <- FitRBM(
  seuratObject = pbmc,
  hiddenFactors = "CellType",
  family = "zinb",
  cd_k = 10, # More Gibbs steps
  n_epochs = 50, # More epochs
  learning_rate = 0.005,
  batch_size = 64,
  persistent = TRUE, # Persistent Contrastive Divergence
  momentum = 0.9,
  weight_decay = 0.001,
  parallel = TRUE
)

# Check training convergence
plot(rbm_trained$training_error,
  type = "l",
  xlab = "Epoch", ylab = "Reconstruction Error",
  main = "RBM Training Convergence"
)
```

You can use multiple metadata factors as hidden units:

```{r eval=FALSE}
# Fit RBM with multiple hidden factors
rbm_multi <- FitRBM(
  seuratObject = pbmc,
  visibleFeatures = NULL,
  hiddenFactors = c("CellType", "Treatment", "Batch"),
  family = "zinb",
  cd_k = 5,
  n_epochs = 20,
  parallel = TRUE
)
```


### 3. Visualize Results

#### Heatmap of Partial Correlations

Visualize the partial correlation structure among features:

```{r eval=FALSE}
# Plot partial correlations for all features
heatmap_pcor <- PlotRBMHeatmap(
  rbmObject = rbm,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  color_palette = "RdBu",
  title = "Partial Correlations: Gene-Gene Relationships"
)

# Draw the heatmap
draw(heatmap_pcor)
```

Plot partial correlations for specific features of interest:

```{r eval=FALSE}
# Select marker genes
marker_genes <- c(
  "CD3D", "CD3E", "CD8A", "CD4", # T cells
  "CD19", "MS4A1", "CD79A", # B cells
  "CD14", "LYZ", "FCGR3A"
) # Monocytes

heatmap_markers <- PlotRBMHeatmap(
  rbmObject = rbm,
  features = marker_genes,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  color_palette = "RdBu",
  title = "Partial Correlations: Cell Type Markers"
)

draw(heatmap_markers)
```

Try different color palettes:

```{r eval=FALSE}
# Viridis color scheme
heatmap_viridis <- PlotRBMHeatmap(rbm, color_palette = "viridis")

# Plasma color scheme
heatmap_plasma <- PlotRBMHeatmap(rbm, color_palette = "plasma")

# Red-Yellow-Blue diverging
heatmap_rylbu <- PlotRBMHeatmap(rbm, color_palette = "RdYlBu")
```

#### Heatmap of RBM Weights

Visualize the learned connections from features to hidden factors:

```{r eval=FALSE}
# Plot all weights
heatmap_weights <- PlotRBMWeights(
  rbmObject = rbm,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  title = "RBM Weights: Features to Cell Types"
)

draw(heatmap_weights)
```

Plot weights for specific features and factors:

```{r eval=FALSE}
# Focus on T cell and B cell markers
heatmap_subset <- PlotRBMWeights(
  rbmObject = rbm_multi,
  features = marker_genes,
  factors = "CellType",
  cluster_rows = TRUE,
  title = "Cell Type Marker Weights"
)

draw(heatmap_subset)
```

### 4. Make Predictions

Use the fitted RBM to predict hidden factor values from expression data:

```{r eval=FALSE}
# Extract expression for new cells
new_expr <- GetAssayData(pbmc_new, assay = "RNA", slot = "counts")

# Predict hidden layer activations
predictions <- predict(rbm, newdata = new_expr, type = "activation")

# The predictions are continuous activation values for each hidden factor
head(predictions)
```

### 5. Reconstruct Expression

Reconstruct expression patterns from hidden factors:

```{r eval=FALSE}
# Reconstruct from training data metadata
reconstructed <- ReconstructRBM(rbm)

# Reconstruct from specific hidden values
# Example: What expression pattern is associated with cell type 1?
hidden_vals <- matrix(c(1, 0, 0), nrow = 1, ncol = 3)
colnames(hidden_vals) <- c("TypeA", "TypeB", "TypeC")

reconstructed_specific <- ReconstructRBM(rbm, hidden = hidden_vals)
```

## Advanced Usage

### Parallel Processing with progressr

For large datasets, enable parallel processing and progress reporting:

```{r eval=FALSE}
library(progressr)

# Configure progressr handlers
handlers(global = TRUE)
handlers("progress")

# Fit RBM with parallel processing
with_progress({
  rbm_parallel <- FitRBM(
    seuratObject = pbmc_large,
    hiddenFactors = c("CellType", "Sample"),
    family = "zinb",
    parallel = TRUE,
    numWorkers = 8,
    progressr = TRUE,
    verbose = TRUE
  )
})
```

### Using Different Distribution Families

Compare results across different families:

```{r eval=FALSE}
# Zero-inflated negative binomial (recommended for scRNA-seq)
rbm_zinb <- FitRBM(pbmc, hiddenFactors = "CellType", family = "zinb")

# Negative binomial
rbm_nb <- FitRBM(pbmc, hiddenFactors = "CellType", family = "nb")

# Poisson
rbm_poisson <- FitRBM(pbmc, hiddenFactors = "CellType", family = "poisson")

# Gaussian (for normalized data)
pbmc_norm <- NormalizeData(pbmc)
expr_norm <- GetAssayData(pbmc_norm, slot = "data")
rbm_gaussian <- FitRBM(pbmc_norm, hiddenFactors = "CellType", family = "gaussian")
```

### Including Metadata Covariates

Include additional metadata as covariates when estimating partial correlations:

```{r eval=FALSE}
# Extract metadata
metadata <- pbmc@meta.data[, c("nCount_RNA", "percent.mt")]

# Estimate partial correlations with covariates
pcor_with_covariates <- EstimatePartialCorrelations(
  expressionMatrix = expr_matrix,
  metadata = metadata,
  family = "zinb",
  parallel = TRUE
)
```

### Subset to Specific Features

Focus analysis on genes of interest:

```{r eval=FALSE}
# Define gene sets
tcell_genes <- c("CD3D", "CD3E", "CD8A", "CD4", "IL7R")
bcell_genes <- c("CD19", "MS4A1", "CD79A", "CD79B")

# Fit RBM with specific features
rbm_subset <- FitRBM(
  seuratObject = pbmc,
  visibleFeatures = c(tcell_genes, bcell_genes),
  hiddenFactors = "CellType",
  family = "zinb"
)
```

## Interpretation

### Partial Correlations

Partial correlations represent the direct relationship between two features after conditioning on all other features:

- **Positive values** (red): Features that co-vary together
- **Negative values** (blue): Features with opposing patterns
- **Values near zero** (white): Weak or no direct relationship

### RBM Weights

The weight matrix shows how features connect to hidden factors:

- **Positive weights**: Feature activates with the factor
- **Negative weights**: Feature is suppressed by the factor
- **Large magnitude**: Strong connection between feature and factor

### Clustering Patterns

When clustering is enabled in heatmaps, groups of features with similar patterns emerge:

- **Feature clusters**: Genes with similar partial correlation patterns (likely co-regulated)
- **Factor associations**: Features that respond similarly to metadata factors

## Performance Tips

1. **Start small**: Test with a subset of features before analyzing the full dataset
2. **Use parallel processing**: Set `parallel = TRUE` for large datasets
3. **Filter features**: Use `minNonZero` to exclude sparse features
4. **Monitor progress**: Enable `progressr = TRUE` to track computation
5. **Choose appropriate family**: Use "zinb" for count data, "gaussian" for normalized data

## Complete Example

Here's a complete workflow from start to finish using the recommended two-step approach:

```{r eval=FALSE}
library(PreGraphModeling)
library(Seurat)
library(ComplexHeatmap)
library(progressr)

# Load data
pbmc <- readRDS("pbmc3k.rds")

# Step 1: Pre-compute partial correlations
# This is the most expensive step - do it once and reuse
handlers(global = TRUE)
with_progress({
  pcor <- EstimatePartialCorrelationsFromSeurat(
    seuratObject = pbmc,
    family = "zinb",
    minNonZero = 10,
    parallel = TRUE,
    numWorkers = 4,
    verbose = TRUE
  )
})

# Inspect the partial correlations
print(pcor)

# Save partial correlations for later reuse
saveRDS(pcor, "pbmc_partial_correlations.rds")

# Step 2: Fit RBM with the pre-computed partial correlations
with_progress({
  rbm <- FitRBM(
    seuratObject = pbmc,
    hiddenFactors = c("CellType", "Phase"),
    partialCorrelations = pcor, # Use pre-computed!
    cd_k = 5,
    n_epochs = 20,
    parallel = TRUE,
    numWorkers = 4,
    verbose = TRUE
  )
})

# View summary
summary(rbm)

# Fit another RBM with different hidden factors (no recomputation!)
rbm_celltype <- FitRBM(
  seuratObject = pbmc,
  hiddenFactors = "CellType",
  partialCorrelations = pcor,
  cd_k = 5,
  n_epochs = 20
)

# Visualize partial correlations
heatmap_pcor <- PlotRBMHeatmap(
  rbmObject = rbm,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  color_palette = "RdBu",
  title = "Gene-Gene Partial Correlations"
)
draw(heatmap_pcor)

# Visualize weights
heatmap_weights <- PlotRBMWeights(
  rbmObject = rbm,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  title = "Feature-Factor Connections"
)
draw(heatmap_weights)

# Save results
saveRDS(rbm, "pbmc_rbm_model.rds")

# Export partial correlation matrix
write.csv(pcor$partial_cor, "partial_correlations.csv")

# Export weights (from first hidden layer)
write.csv(rbm$weights_per_layer[[1]], "rbm_weights_celltype.csv")
```

## Session Info

```{r eval=FALSE}
sessionInfo()
```

## References

- Restricted Boltzmann Machines provide a probabilistic framework for unsupervised learning
- Partial correlations capture direct relationships while controlling for confounders
- Pseudolikelihood provides a computationally efficient approximation to full likelihood training by conditioning on neighbors
- ComplexHeatmap enables sophisticated visualization of high-dimensional data

## See Also

- `?EstimatePartialCorrelationsFromSeurat` - Compute partial correlations from Seurat object (recommended)
- `?EstimatePartialCorrelations` - Compute partial correlations from expression matrix
- `?FitRBM` - Fit restricted Boltzmann machine to Seurat object
- `?PlotRBMHeatmap` - Visualize partial correlations as heatmap
- `?PlotRBMWeights` - Visualize RBM weights as heatmap
- `?predict.RBM` - Make predictions from fitted RBM
- `?ReconstructRBM` - Reconstruct visible layer from hidden factors
